From DSL to HPC Component-Based Runtime:\\ A Multi-Stencil DSL Case Study

### 2015-07-31 WOLFHPC15/article/component.tex
fix all helene 30_07

%----------------------------------------
\subsection{Overview}
%----------------------------------------

This section describes both the projection of $\Gamma_{tsp}$ to
components, and the remaining needed components to execute a
multi-stencil application. But, first the section gives an overview of
component models, and it details \llc, the component model used in
this work.

%----------------------------------------
\subsection{Component Model Overview}
%----------------------------------------
Component model is an domain of software
engineering~\cite{Szyperski:2002:CSB:515228}, which promotes code
re-use, separations of concerns, and thus maintainability. An
application is made by a set of component; a component being a kind of
black box that interacts only through ports: ports specify the
services provided and required.
%
With respect to high performance computing, some works have also shown
that component models can achieve the needed level of performance, and
scalability while also helping in application
portability~\cite{Bernholdt01052006, bigot:inria-00388508, UCHPC2015}

Many component models exist, each of them with its own specifications
and singularities. In distributed computing, well known component
models are CCM~\cite{corba:omg06} (CORBA Component Model), and
GCM~\cite{Baude} (Grid Component Model); in HPC, there are
CCA~\cite{Bernholdt01052006} (Common Component Architecture), and
\llc~\cite{l2c} (Low Level Component), for example.
%
As this work makes use of \llc, in particular for the experiments, let
describe it in more details.

@@

\llc is a minimalist \texttt{C++} based HPC-oriented component model
where a component extends the concept of class by specifying in its
interfaces the services that it offers ($provide$ ports) and that it
needs, either a single service instance ($use$ ports), or multiple
service instances ($use-multiple$ ports). Services are \texttt{C++}
interfaces. \llc also offer $MPI$ ports that enable components to
share MPI communicators. Components can also have attribute ports to
be configured.
%
As illustrated in Figures~\ref{fig:ports}, a $provide$ port is
represented with a white circle on the left of the component, a $use$
port with a black circle, a $use-multiple$ port by a black circle with
a white $m$ in it, both on the right of the component. MPI port are
connected with a black rectangle.

@@

\begin{figure}[t]
a)\hspace{2mm} \begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (C) at (0,0) {$C_0$};
   \node[provide] (p) at (-1.5,0) {};
   \node[use] (u) at (1.5,0) {};
   \node[provide,right=1.5cm of u] (p1) {};
   \node[component,right=1.5cm of p1] (C1) {$C_1$};
   \node[use,right=1.5cm of C1] (um) {$m$};

  \path[-]
    (p) edge node {$p$} (C)
    (C.east) edge node {$u$} (u)
    (C1)  edge node {$v$} (um)
    (p1) edge node {$q$} (C1);
\end{tikzpicture}
\\\\
b)\hspace{2mm}\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (C) at (0,0) {$C_0$};
   \node[provide] (p) at (-1.5,0) {};
   \node[use] (u) at (1.5,0) {};
   \node[component,right=1.5 of u] (C1) {$C_1$};
   \node[use,right=1.5 of C1] (um) {$m$};

  \path[-]
    (p) edge node {$p$} (C)
    (C) edge node {$u$} (u)
    (C1)  edge node {$v$} (um)
    (u) edge node {$q$} (C1);
\end{tikzpicture}
\\\\
c) \hspace{2mm}\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (C) at (0,0) {$C_2$};
   \node[provide] (p) at (-1.5,0) {};
   \node[mpi] (u) at (1.5,0) {};
   \node[component,right=1.5 of u] (C1) {$C_3$};

  \path[-]
    (p) edge node {$r$} (C)
    (C) edge node {$m_1$} (u)
    (u) edge node {$m_2$} (C1);
\end{tikzpicture}
\\\\
\caption{Example components and ports representation. a) Component $C_0$ has a provide port ($p$) and a use port ($u$); Component $C_1$ has also a provide port ($p'$) but a use multiple port ($u'$). b) A use port is connected to a (compatible) provide port. c) Component $C_2$ and $C_3$ shares an MPI communicator.}
\label{fig:ports}
\end{figure}

@@

Figure~\ref{fig:mscac:assembly} has presented an overview of the component
assembly that is executed. As discusses in Section~\ref{sect:mscac},
there are different types of components. Components~\texttt{Driver},
\texttt{DriverApp}, and \texttt{Time} are always present; they do not
depend on the input but for the configuration of an attribute port of
Component~\texttt{Time} with the number of iterations. There is also a
single Component~\texttt{DDS} as in the current version only one mesh
type is supported. All these components are provided with the
compiler; dumping this part of the assembly is straightforward.

@@

The last part of the assembly is the computations. It contains actual
computations in kernel components, denoted K, that have to be provided
to the compiler. A kernel component encapsulates a computations. To
access data, they are connected by the compiler to the adequate
\texttt{Data} components; their data ports name must respect the names
given in the \texttt{data} section of MSL. We denote the fact that they may have
several use port by adding a star on the use port:
\begin{center}
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (k) at (0,0) {$K$};
   \node[provide] (p) at (-1.5,0) {};
   \node[use] (u) at (1.5,0) {};
   \node[right=0.2 of u] (star) {*};

  \path[-]
    (p) edge node {} (k)
    (k) edge node {} (u);
\end{tikzpicture}
\end{center}

@@

The computation part of the assembly also contains the static schedule
of the computations, \ie $\Gamma_{tsp}$. This is achieved by using
three specific components for managing the $P$, $S$, and $sync$
operations of $\Gamma_{tsp}$ as explained hereafter.

%----------------------------------------
\subsection{Control components}
%----------------------------------------
The series-parallel tree decomposition $\Gamma_{tsp}$ represents the
control of the dependencies of the simulation. As a result to be able
to dump it to a component assembly it is needed to introduce what can
be called \emph{control components}. Those components can be used for
any case of simulation which increases code reuse between
simulations. A control component is composed of a single
\emph{provide} port linked to a single execution method, most of the
time called the \emph{go} method. We introduce three types of control
components, whole graphical representation is given in Figure~\ref{fig:ctrlcomponents}.

@@

\item[Synchronization component (SYNC)]. It is the direct representation to an update leaf of $\Gamma_{tsp}$. The role of this component is to call the synchronization of a given data. It offers an \emph{use} port to be connected to the data to update.

@@

%----------------------------------------
\subsection{Dump to a component assembly}
%----------------------------------------
From $\Gamma_{tsp}$ and using the control components described above,
a direct dump can be done to a component assembly for an hybrid (data
and task) parallelization of a simulation. However, it is also
possible to dump to a data parallelization only. In such a case the
computation of $\Gamma_{data}$ is enough to generate the component
assembly. Actually a single SEQ component is thus needed and this
component is linked to all computations and synchronizations of
$\Gamma_{data}$ directly.

Figure~\ref{fig:tsp:assembly} displays the assembly part corresponding to
$\Gamma_{tsp}$ of Figure~\ref{fig:tsp}. In this figure, the ports
linked to data (use and use-multiple ports of SYNC and K) are
represented but are not connected. Moreover, each computation
component is an instance of the component K presented before but takes
the identification name of the computation in the MSL file.

>>>>

%----------------------------------------
%\subsection{Overview}
%----------------------------------------
This section describes both the projection of $\Gamma_{tsp}$ to
components, and the remaining needed components to execute a
multi-stencil application. Before those details the section gives an overview of
component models, and introduce the concepts of the Low Level Component Model \llc, which is used in this paper.

%----------------------------------------
\subsection{Component Model Overview}
%----------------------------------------
Component model is a domain of software
engineering~\cite{Szyperski:2002:CSB:515228}, which promotes code
re-use, separations of concerns, and thus maintainability. An
application is made of a set of components; a component being a kind of
black box representing an independent functionnality of the application, and which interacts only through its ports. A port specifies the services provided and required by the component.
%
With respect to high performance computing, some works have also shown
that component models can achieve the needed level of performance, and
scalability while also helping in application
portability~\cite{Bernholdt01052006, bigot:inria-00388508, UCHPC2015}

Many component models exist, each of them with its own specifications
and singularities. In distributed computing, well known component
models are CCM~\cite{corba:omg06} (CORBA Component Model), and
GCM~\cite{Baude} (Grid Component Model); in HPC, there are
CCA~\cite{Bernholdt01052006} (Common Component Architecture), and
\llc~\cite{l2c} (Low Level Component), for example.
%
As this work makes use of \llc, in particular for the experiments, let
introduce concepts of this component model with more details.

@@

\llc is a minimalist \texttt{C++} based HPC-oriented component model
where a component extends the concept of class by specifying in its
interfaces the services that it offers ($provide$ ports) and that it
needs, either a single service instance ($use$ ports), or multiple
service instances ($use-multiple$ ports). Services are \texttt{C++}
interfaces. \llc also offers $MPI$ ports that enable components to
share MPI communicators. Components can also have attribute ports to
be configured.
%
As illustrated in Figures~\ref{fig:ports}, a $provide$ port is
represented with a white circle, a $use$
port with a black circle, a $use-multiple$ port by a black circle with
a white $m$ in it. MPI port are
connected with a black rectangle.

@@

\begin{figure}[t]
\begin{center}
\subfloat[][\label{fig:2comp}]{
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (C) at (0,0) {$C_0$};
   \node[provide] (p) at (-1.5,0) {};
   \node[use] (u) at (1.5,0) {};
   \node[provide,right=1.5cm of u] (p1) {};
   \node[component,right=1.5cm of p1] (C1) {$C_1$};
   \node[use,right=1.5cm of C1] (um) {$m$};

  \path[-]
    (p) edge node {$p$} (C)
    (C.east) edge node {$u$} (u)
    (C1)  edge node {$v$} (um)
    (p1) edge node {$q$} (C1);
\end{tikzpicture}
}
\subfloat[][\label{fig:ass}]{
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (C) at (0,0) {$c_0$};
   \node[provide] (p) at (-1.5,0) {};
   \node[use] (u) at (1.5,0) {};
   \node[component,right=1.5 of u] (C1) {$c_1$};
   \node[use,right=1.5 of C1] (um) {$m$};

  \path[-]
    (p) edge node {$p$} (C)
    (C) edge node {$u$} (u)
    (C1)  edge node {$v$} (um)
    (u) edge node {$q$} (C1);
\end{tikzpicture}
}
\subfloat[][\label{fig:mpi}]{
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (C) at (0,0) {$c_2$};
   \node[provide] (p) at (-1.5,0) {};
   \node[mpi] (u) at (1.5,0) {};
   \node[component,right=1.5 of u] (C1) {$c_3$};

  \path[-]
    (p) edge node {$r$} (C)
    (C) edge node {$m_1$} (u)
    (u) edge node {$m_2$} (C1);
\end{tikzpicture}
}
\caption{Example of components and their ports representation. a) Component $c_0$ has a provide port ($p$) and a use port ($u$); Component $c_1$ has also a provide port ($q$) but also a use multiple port ($v$). b) A use port is connected to a (compatible) provide port. c) Component $c_2$ and $c_3$ shares an MPI communicator.}
\label{fig:ports}
\end{figure}

@@

Figure~\ref{fig:mscac:assembly} gives an overview of the back-end component
assembly that is executed. As discusses in Section~\ref{sect:mscac},
there are different types of components. Components~\texttt{Driver},
\texttt{DriverApp}, and \texttt{Time} are always instanciated in the final component assembly; the input MSL file is not used by them except the component \texttt{Time} which uses the terminal \texttt{iteration} of the grammar. There is also a
single Component~\texttt{DDS} (the current version handles a single mesh
type). All these components are provided with the
compiler; dumping this part of the assembly is straightforward.

@@

The last part of the assembly is the computation part. Each computation kernel is embedded in a kernel component, denoted K, which have to be written by the user and provided
to the compiler. A kernel component encapsulates a computations. To
access data, a kernel component is connected, by the compiler, to the adequate
\texttt{Data} components such that their ports name respect the names
given in the \texttt{data} section of MSL. We denote the fact that they may have
several use port by adding a star on the use port, as illustrated in Figure~\ref{fig:k}.
\begin{figure}
\begin{center}
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[component] (k) at (0,0) {$K$};
   \node[provide] (p) at (-1.5,0) {};
   \node[use] (u) at (1.5,0) {};
   \node[right=0.2 of u] (star) {*};

  \path[-]
    (p) edge node {} (k)
    (k) edge node {} (u);
\end{tikzpicture}
\end{center}
\caption{Kernel component.}
\label{fig:k}
\end{figure}

@@

The rest of the computation part of the assembly contains the static parallel schedule
computed by MSC, \ie $\Gamma_{tsp}$. This is achieved through the use of
three specific components which manage the $P$, $S$, and $sync$
operations of $\Gamma_{tsp}$ as explained hereafter.

%----------------------------------------
\subsection{Control Components}
%----------------------------------------
The series-parallel tree decomposition $\Gamma_{tsp}$ represents the
control of the dependencies of the simulation. As a result to be able
to dump it to a component assembly it is needed to introduce what can
be called \emph{control components}. Those components can be used for
any case of simulation which increases code reuse between
simulations. A control component is composed of a single
\emph{provide} port linked to a single execution method, most of the
time called the \emph{go} method. We introduce three types of control
components represented in Figure~\ref{fig:ctrlcomponents}.

@@

\item[Synchronization component (SYNC)]. It is the direct representation to an update leaf of $\Gamma_{tsp}$. The role of this component is to call the synchronization of a given data. It offers a \emph{use} port to be connected to the data to update.

@@

%----------------------------------------
\subsection{Dump To Component Assembly}
%----------------------------------------
From $\Gamma_{tsp}$ and using the control components described above,
a direct dump can be done to a component assembly for an hybrid (data
and task) parallelization of a simulation. 
\fix{HC: ce qui suit est a changer avec la fusion du code de plusieurs composants}
However, it is also
possible to dump to a data parallelization only. In such a case the
computation of $\Gamma_{sync}$ is enough to generate the component
assembly. Actually a single SEQ component is thus needed and this
component is linked to all computations and synchronizations of
$\Gamma_{sync}$ directly.

Figure~\ref{fig:tsp:assembly} displays the assembly part corresponding to
$\Gamma_{tsp}$ of Figure~\ref{fig:tsp}. In this figure, the ports
linked to data (use and use-multiple ports of SYNC and K) are
represented but are not connected. Moreover, each computation
component is an instance of the component K presented before but using
the identification name of the computation in the MSL file.

### 2015-07-31 WOLFHPC15/article/concept.tex
fix all helene 30_07

%----------------------------------------
\subsection{Stencil codes}
\label{sect:stencil}
%----------------------------------------
To numerically solve a set of PDEs, iterative methods (finite difference, finite volume, finite element methods etc.) are frequently used to approximate the solution through a discretized (step by step) process. Thus, the continuous time and space domains are discretized so that a set of numerical computations are iteratively (time discretization) applied onto a mesh (space discretization). In other words, in a mesh-based numerical simulation, the PDEs are transformed to a set of numerical computations applied at each time step on elements of the discretized space domain (the mesh).
This paper focuses on one specific category of such numerical schemes based on \textit{stencil kernels}~\cite{spaaTangCKLL11}.
This section introduces some definitions related to those.

A \emph{mesh} is the discretization of a physical domain. It is a connected undirected graph without bridges (an edge is a bridge if its removal results in two disconnected graphs), where nodes and edges are linked to form cells (closure). An example of structured mesh is illustrated in Figure~\ref{fig:mesh}. Each cell contains four nodes and four edges. \emph{Mesh entities} are elements of the mesh, such as the center of cells, edges or nodes. In Figure~\ref{fig:mesh}, two mesh entities are illustrated, the center of the cells (named \texttt{Cells}, in red) and edges on the horizontal axis (named \texttt{Edgex}, in blue on left and right of each cell). A \emph{data} is a simulated quantity on which computations are performed. Each data is mapped onto mesh entities. For example, in Figure~\ref{fig:ex1}, data $A$ and $B$ are mapped onto the center of cells, while, in Figure~\ref{fig:ex2}, $C$ is mapped onto the edges.
\todo[inline]{JB: structured/unstructured pas defini, pas de notion de coordonees des elements du maillage dans l'espace reel?}

@@

A \emph{stencil kernel} computes the value of one data or a subpart of it (the kernel \emph{computation domain}) using a \emph{numerical expression} taking as input on one or more data.
For example, the stencil kernel illustrated in Figure~\ref{fig:ex1} computes $A$ using $B$, while the one in Figure~\ref{fig:ex2} computes $A$ using $C$. Thus, a stencil kernel is defined by its numerical expression, a set of input data (only one in the examples), and its unique output data, the result. 
The computation domain is a subset of the mesh entities on which the output data is mapped. For example, in the first stencil, the computation of $A$ is performed on plain elements, while dotted elements are not computed. On the other hand, on the second computation, the computation domain of $A$ contains all the mesh entities on which it is mapped. This computation domain defines the elements over which the space loop iterates.

The numerical expression of a stencil kernel has the particularity to compute each element of the result independently using only elements of the inputs whose coordinates are close to that of the result element.
The elements accessed form a \emph{neighborhood} of the output element coordinate known as the \emph{stencil shape}. For example, the stencil shape of the first computation in Figure~\ref{fig:ex1} contains direct neighbors on the right, left, top and bottom. Sometimes, the neighborhood can also access different mesh entities, as for example in Figure~\ref{fig:ex2}. Actually, in this computation, the neighborhood contains edges on the left and on the right of a cell. As an example, the numerical expression of the first example could be:
\begin{equation*} 
A(x,y) = B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1),
\end{equation*}
and the numerical expression of the second kernel could be:
\begin{equation*} 
A(x,y) = A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

As can be seen from these definitions, stencil kernels implementations share many properties from an algorithmic point of view that can be used to apply well known optimization and parallelization strategies.
Many solutions (languages or libraries) thus propose to ease their programming by producing optimized and parallelized code from a simple description of the local computation to apply on each element.
For example, Liszt~\cite{DeVito2011LDS} and OP2~\cite{Giles2011} are domain specific languages to define a stencil code on unstructured meshes, while Pochoir~\cite{spaaTangCKLL11} and PATUS~\cite{citeulike12258902} address structured meshes, respectively proposing an optimized cache tiling technique, and a separation of concerns between parallelization strategies and computation descriptions.
\todo[inline]{JB: a virer?}

To resume, concepts used in this paper to define a stencil kernel are \emph{mesh}, \emph{mesh entities}, \emph{data}, \emph{computation domain}, \emph{numerical expression} and \emph{stencil shape}. 
If stencil kernels have been studied a lot, the formalization of real overall mesh-based numerical simulations is a rather new contribution. Actually, paying attention to complex numerical simulations, it appears that most of them are composed of more than one stencil kernel, with one or more stencil shapes and of additional local computations. For example, we would like to formalize and parallelize a numerical simulation which chains stencil kernels of Figures~\ref{fig:ex1} and~\ref{fig:ex2}. The next section formalizes concepts of \emph{stencil kernel}, \emph{local computation} and what we call a \emph{multi-stencil program}.

>>>>

%----------------------------------------
\subsection{Stencil Kernels}
\label{sect:stencil}
%----------------------------------------
To numerically solve a set of PDEs, iterative methods (finite difference, finite volume, finite element methods etc.) are frequently used to approximate the solution through a discretized (step by step) process. Thus, the continuous time and space domains are discretized so that a set of numerical computations are iteratively (time discretization) applied onto a mesh (space discretization). In other words, in a mesh-based numerical simulation, the PDEs are transformed to a set of numerical computations applied at each time step on elements of the discretized space domain (the mesh).
This paper focuses on one specific category of such numerical schemes based on \textit{stencil kernels}~\cite{spaaTangCKLL11}, also called in applied mathematics explicit numerical schemes.
This section introduces some definitions related to stencil kernels.

A \emph{mesh} is the discretization of a physical domain. It is a connected undirected graph without bridges (an edge is a bridge if its removal results in two disconnected graphs), where nodes and edges are linked to form cells (closure). Most of the time, nodes of a mesh are linked to coordinates in the real space domain. An example of structured mesh is illustrated in Figure~\ref{fig:mesh}. Each cell contains four nodes and four edges. \emph{Mesh entities} are elements of the mesh, such as the center of cells, edges or nodes. In Figure~\ref{fig:mesh}, two mesh entities are illustrated, the center of the cells (named \texttt{Cells}, in red) and edges on the horizontal axis (named \texttt{Edgex}, in blue on left and right of each cell). A \emph{data} is a simulated quantity on which computations are performed. Each data is mapped onto mesh entities. For example, in Figure~\ref{fig:ex1}, data $A$ and $B$ are mapped onto the center of cells, while, in Figure~\ref{fig:ex2}, $C$ is mapped onto the edges.
%\todo[inline]{JB: structured/unstructured pas defini, pas de notion de coordonees des elements du maillage dans l'espace reel?}

@@

A \emph{stencil kernel} computes the value of one data or a subpart of it (the kernel \emph{computation domain}) using a \emph{numerical expression} which takes as input one or more data.
For example, the stencil kernel illustrated in Figure~\ref{fig:ex1} computes $A$ using $B$, while the one in Figure~\ref{fig:ex2} computes $A$ using $C$. Thus, a stencil kernel is defined by its numerical expression, a set of input data (only one in the examples), and its unique output data, the result. 
The \emph{computation domain} is a subset of the mesh entities on which the output data is mapped. For example, in the first stencil, the computation of $A$ is performed on elements represented with full lines, while dotted elements are not computed. On the other hand, on the second computation, the computation domain of $A$ contains all the mesh entities on which it is mapped. This computation domain defines the elements over which the space loop iterates.

The numerical expression of a stencil kernel has the particularity to compute each element of the result independently using some elements of the inputs in a given neighborhood.
Accessed elements form a \emph{neighborhood} of the output known as the \emph{stencil shape}. For example, the stencil shape of the first computation in Figure~\ref{fig:ex1} contains direct neighbors on the right, left, top and bottom. Sometimes, the neighborhood can also access different mesh entities, as for example in Figure~\ref{fig:ex2}. Actually, in this computation, the neighborhood contains edges on the left and on the right of a cell. As an example, the numerical expression of the first example could be:
\begin{equation*} 
A(x,y) = B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1),
\end{equation*}
and the numerical expression of the second kernel could be:
\begin{equation*} 
A(x,y) = A(x,y)+C(x1,y1)+C(x1+1,y1).
\end{equation*}

As it can be seen from those definitions, stencil kernels implementations share many properties from an algorithmic point of view that can be used to apply well known optimization and parallelization strategies.
Many solutions (languages or libraries) thus propose to ease their programming by producing optimized and parallelized code from a simple description of the local computation to apply on each element.
For example, Liszt~\cite{DeVito2011LDS} and OP2~\cite{Giles2011} are domain specific languages to define a stencil code on unstructured meshes, while Pochoir~\cite{spaaTangCKLL11} and PATUS~\cite{citeulike12258902} address structured meshes, respectively proposing an optimized cache tiling technique, and a separation of concerns between parallelization strategies and computation descriptions.
\todo[inline]{JB et HC: etat de l'art a mettre ailleurs ?}

%To resume, concepts used in this paper to define a stencil kernel are \emph{mesh}, \emph{mesh entities}, \emph{data}, \emph{computation domain}, \emph{numerical expression} and \emph{stencil shape}. 
If stencil kernels have been studied a lot, the formalization of real overall mesh-based numerical simulations is a rather new contribution. Actually, paying attention to complex numerical simulations, it appears that most of them are composed of more than one stencil kernel, with one or more stencil shapes and of additional local computations. For example, we would like to formalize and parallelize a numerical simulation which chains stencil kernels of Figures~\ref{fig:ex1} and~\ref{fig:ex2}. The next section formalizes concepts of \emph{stencil kernel}, \emph{local computation} and what we call a \emph{multi-stencil program}.

### 2015-07-31 WOLFHPC15/article/evaluation.tex
fix evaluation helene 30_07
fix all helene 30_07

One can notice that the transformation of $\Gamma_{dep}$ to a minimal series-parallel graph is the longest step of MSC, because of the removal of the forbidden shapes in the graph. Actually, the number of forbidden shapes removed in $\Gamma_{dep}$ is not counted, because the algorithm use a general solution instead of finding each forbidden shape, but it seems that many of them appeared. The forbidden shape, called \emph{N-shape}, is represented in Figure~\ref{fig:n}. This shape is forbidden in a minimal series-parallel graph as it is not possible to exactly express it using sequences and parallel sections. To remove a \emph{N-shape} an additionnal dependency constraint is added as illustrated in Figure~\ref{fig:over}.

\begin{figure}[h!]
\begin{center}
\subfloat[][The forbidden $N$ shape.\label{fig:n}]{
\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[] (a) at (0,0) {$c_0$};
   \node[] (b) at (2,0) {$c_1$};
   \node[] (c) at (0,-1) {$c_2$};
   \node[] (d) at (2,-1) {$c_3$};

  \path[->]
    (a) edge node {} (b)
    (c) edge node {} (b)
        edge node {} (d);
  \end{tikzpicture}
  }
  \hspace{25pt}
  \subfloat[][Over-constraint on the forbidden $N$ shape.\label{fig:over}]{
	\begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
   \node[] (c0) at (0,0) {$c_0$};
   \node[] (bc0) at (-1,0) {};
   \node[] (c1) at (2,0) {$c_1$};
   \node[] (c2) at (0,-1) {$c_2$};
   \node[] (bc2) at (-1,-1) {};
   \node[] (c3) at (2,-1) {$c_3$};

  \path[->]
    (bc0) edge [dotted] node {} (c0)
    (bc2) edge [dotted] node {} (c2)
    (c0) edge node {} (c1)
          edge [dashed] node [swap] {} (c3)
    (c2) edge node {} (c1)
        edge node {} (c3);
  \end{tikzpicture}
}
  \caption{Forbidden $N$ subgraph shape for a DAG to be minimal series-parallel.}
  \label{fig:forbidden}
\end{center}
\end{figure}

The fact that many \emph{N-shape} are removed in $\Gamma_{dep}$ shows that the creation of a static shedule of tasks may not be the best solution for complex simulations. Actually, if a \emph{N-shape} cannot be represented by sequences and parallel sections, this shape can perfectly be handled by a dynamic scheduler or by the direct static representation of $\Gamma_{dep}$ where tasks wait for their input before starting.

>>>>

\fix{HC 2 all : Je ne sais pas si ce qui suit a sa place dans cet article.}

One can notice that the transformation of $\Gamma_{dep}$ to a minimal series-parallel graph is the longest step of MSC, because of the removal of the forbidden shapes in the graph. Actually, the number of forbidden shapes removed in $\Gamma_{dep}$ is not counted, because the algorithm use a general solution instead of finding each forbidden shape, but it seems that many of them appeared. The \emph{N-shape}, represented in Figure~\ref{fig:n} is forbidden in a minimal series-parallel graph as it is not possible to exactly express it using sequences and parallel sections.
The fact that many \emph{N-shape} are removed in $\Gamma_{dep}$ shows that the creation of a static shedule of tasks may not be the best solution for complex simulations. Actually, if a \emph{N-shape} cannot be represented by sequences and parallel sections, this shape can perfectly be handled by a dynamic scheduler or by the direct static representation of $\Gamma_{dep}$ where tasks wait for their input before starting.

### 2015-07-31 WOLFHPC15/article/introduction.tex
fix all helene 30_07

For this reason, solutions to ease the design and implementation of DSLs have recently appeared~\cite{Fernandez:2014:DFL:2691166.2691168}. The main idea of those solutions is to propose a single DSL to write new DSLs. In this paper, we study a first contribution to another approach to improve maintainability and portability in DSLs. This approach proposes a transformation from a DSL to a component-based runtime. Component-based Software Engineering have proved many times good properties for code re-use, separation of concerns, maintainability and productivity of codes. For this reason, we think that mixing DSLs and component-based runtimes could enable those properties, by inheritance, in DSLs.
To evaluate this approach, a proper new DSL for \emph{multi-stencil} programs, and its compiler are presented in this paper.
%\fix{cp2cp+hc: to dev benefits of components.}

Many solutions to automatically optimize and parallelize partial differential equations (PDEs) solvers have been proposed in the literature, either as specific libraries~\cite{petsc-efficient,Trilinos-Overview,CPE:CPE3494} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil kernel.
However, a real case numerical simulation is most of the time not composed of a single stencil kernel, but of a set of stencil kernels and a set of additional local auxiliary computations. The DSL proposed in this paper, called MSL, is a DSL to generate a parallel component-based structure of an overall numerical simulation, that we call a \emph{multi-stencil program}. As it will be detailed in the related work, this is complementary to the optimization and parallelization of a single stencil kernel.
%\fix{JB: j'utiliserais bien le terme "kernel" ou "stencil kernel" pour ce que tu apelles "stencil" et "stencil" tout court pour ce que tu apelles "stencil shape"}

The rest of this paper is organized as follows. First, the concepts of \emph{stencil} and \emph{multi-stencil programs} are formalized in Section~\ref{sect:concept}. Section~\ref{sect:mscac} gives an overview of the solution presented in this paper while Section~\ref{sect:msmsc} details the MSL language and its compiler. Section~\ref{sect:component} introduces component models and describes the proposed component-based runtime and the automatic generation of its parallel computation part.
%The end of Section~\ref{sect:component} also discusses on interest of this new approach.
Section~\ref{sect:eval} focuses on a real-case simulation, solving the Shallow-water equations, and analyzes some performance results. Sections~\ref{sect:related} states on related work, and finally Section~\ref{sect:conclusion} concludes and gives some perspectives.

>>>>

For this reason, solutions to ease the design and implementation of DSLs have recently appeared~\cite{Fernandez:2014:DFL:2691166.2691168}. The main idea of those solutions is to propose a single DSL or a framework to write new DSLs. In this paper, we study a first contribution to another approach to improve maintainability and portability in DSLs. This approach proposes a transformation from a DSL to a component-based runtime. Component-based Software Engineering have proved many times good properties for code re-use, separation of concerns, maintainability and productivity of codes. For this reason, we think that mixing DSLs and component-based runtimes could enable those properties, by inheritance, in DSLs.
To evaluate this approach, a proper new DSL for \emph{multi-stencil} programs, and its compiler are presented in this paper.
%\fix{cp2cp+hc: to dev benefits of components.}

Many solutions to automatically optimize and parallelize partial differential equations (PDEs) solvers have been proposed in the literature, either as specific libraries~\cite{petsc-efficient,Trilinos-Overview,CPE:CPE3494} or as DSLs~\cite{spaaTangCKLL11,citeulike12258902,Giles2011,DeVito2011LDS}. Most of the time, DSLs focus on the parallelization and the optimization of a single numerical computation, also called a stencil kernel.
However, a real case numerical simulation is most of the time not composed of a single stencil kernel, but of a set of stencil kernels and a set of additional local auxiliary computations. The DSL proposed in this paper, called MSL, is a DSL to generate a parallel component-based structure of an overall numerical simulation, that we call a \emph{multi-stencil program}. As it will be detailed in the related work, this is complementary to the optimization and parallelization of a single stencil kernel.

The rest of this paper is organized as follows. First, the concepts of \emph{stencil kernels} and \emph{multi-stencil programs} are formalized in Section~\ref{sect:concept}. Section~\ref{sect:mscac} gives an overview of the solution presented in this paper while Section~\ref{sect:msmsc} details the MSL language and its compiler. Section~\ref{sect:component} introduces component models and describes the proposed component-based runtime and the automatic generation of its parallel computation part.
%The end of Section~\ref{sect:component} also discusses on interest of this new approach.
Section~\ref{sect:eval} focuses on a real-case simulation, solving the Shallow-water equations, and analyzes some performance results. Sections~\ref{sect:related} states on related work, and finally Section~\ref{sect:conclusion} concludes and gives some perspectives.
